Leadership across cultures and contexts: an exploratory analysis

### 2020-05-03 leadership_across_cultures_contexts.Rmd
Added sex differences and female leadership discussion section.

<!-- MALE BIAS -->
Substantial evidence suggests there is a near universal male-bias in community, political, or group-level leadership across human societies [@low_sex_1992; @whyte_status_1978; @sanday_female_1981], as well as across diverse non-human social species [@smith_obstacles_2018; @tiger_female_1978]. @low_sex_1992 demonstrated in a systematic cross-cultural sample that political leaders of both sexes were present in only seven of 93 societies and in only two societies were female political leaders identified as equally powerful as their male counterparts. Even among these rare cases, powerful female political leaders are still infrequent within the group and women were more likely to hold influential leadership positions within kin groups [@low_sex_1992]. Empirical ethnographic studies have suggested sex-differences in political influence and leadership across human societies including among more egalitarian cultural contexts are more likely driven by ecological and social structural factors, rather than sex (or gender identity) per se, including social capital and access to same-sex allies [@yanca_female_2004] and/or sex differences in the sexual division of labor [@von_rueden_sex_2018]. At least two studies relying on ethnographic fieldwork comparing leaders and non-leaders in relatively small-scale societies revealed sex is not a strong predictor of community influence or elected leadership positions, after accounting for a broad range of demographic, phenotypic, cognitive, social, and other traits derived from the evolutionary theoretical literature [e.g., @garfield_investigating_2019; @von_rueden_sex_2018].

@@

Evolutionary theoretical models and empirical studies often seek to integrate findings within a broad framework of human cultural diversity. The universality of human leadership systems and the incredible diversity among the qualities of leaders across cultures and contexts and the functions leaders serve is difficult to capture. Scholars are then left to haphazardly self-select ethnographic cases or rely on the self-selected reviews of others for generalizations [e.g., @garfield_review_2019; @von_rueden_leadership_2015]. Here, we seek to provide the first empirical and near comprehensive view of the qualities and functions of leaders across human societies from a representative cross-cultural sample of "non-WEIRD" populations. We also provide the first comparative view describing the costs and benefits of leadership for both leaders and followers in an evolutionary biological framework. We also aim to explore how the qualities and functions of leaders vary within populations across social contexts and across populations with different subsistence strategies.

@@

The frequency of leader and follower costs and benefits were estimated similarly to the leader qualities and functions described above. In these data leaders most commonly receivied benefits in the form of greater access to material resources (documented in about 70% of cultures) and increased social status (documented in about 65% of cultures). Mating benefits and receiving social services were also relatively frequent benefits of leadership. In general, the costs tracked the benefits. Leaders were often described as losing their high social status (documented in about 45% of cultures) and greater access to resources (documented in about 40% of cultures). See Figure \@ref(fig:varsupportcb).

Followers on the other hand most commonly benefited from leadership systems through receiving material resources (documented in about 50% of cultures) but otherwise most frequently through social services provided (documented in about 35% of cultures). Mating and fitness benefits were relatively rare for followers. Followers were most frequently described as incuring costs in the form of material resources lost (documented in about 40% of cultures) and in providing social services (documented in about 25% of cultures). See Figure \@ref(fig:varsupportcb).

@@

For subsistence strategy, hunter-gatherers had relatively more evidence for *Polygyny*, *Ledaer-Mating benefits*, *No coercive authority*, *Political appointments* and *Military command* than other subsistence strategies. Pastoralists had more evidence for *Ritual functions*, *Supernatural*, *Aggressiveness* and *High quality spouse*. Hunter-gatherers and pastoralists both had more evidence for *High status*, *Providing counsel/direction*, and *Providing subsistence*. The remaining subsistence strategies generally had relatively less evidence for all variables, except that the mixed strategy had relatively more evidence for *Organize cooperation*, *Distribute resources*, and *Prosocial investment*. See Figure S\@ref(fig:heatmapsubsis).

@@

There was strong patterning of relative evidence for almost all variables by leader sex. Female leaders had relatively more evidence for *High quality spouse*, *Leader - Mating benefits*, *Polygyny*, *High status*, and *Age*. Male leaders had relatively more evidence for *Follower - Social  services benefits*, *Political appointments*, *Prosocial investment*, and *Leader-Social services costs*. Records pertaining to both sexes had relatively more evidence for *Organize cooperation*, *Providing subsistence*, *Ritual functions*, *Bravery*, *Experience/accompished*, and few that overlapped with males, such as *Military command* and *Distributing resources*. See Figure S\@ref(fig:heatmapsex).

There was also strong patterning of relative evidence for almost all variables by group structure. There was relatively more evidence for military traits in military groups, relatively more evidence for ritual and supernatural traits in relgious groups, relatively more evidence for organizing cooperation and providing subsistence in economic groups, and so forth. Kin groups had relatively more evidence for *Provide counsel/direction*, *Prosocial investment*, *Age*, and *High status*. Community and residential groups had relatively low evidence for most variables, except that there was relatively more evidence for *High quality spouses* among residential group leaders. See Figure S\@ref(fig:heatmapgroups).

@@

We then explored relationships between total leader benefits and costs, summed across all leader benefit and cost variables, respectively, and levels of group context. In general, leaders at the highest supracommunity political level had the most evidence for receiving benefits and the least evidence for suffering costs and leaders of military groups has the highest total costs. See Figure \@ref(fig:leadertotalbenefits).

@@

*High status* was the most frequently identified leader quality in the sample, with `r statusPercTxt`% of text records and `r statusPercCult`% of cultures containing evidence that leaders were of high social status. To explore factors that distinguished these records from those with no evidence for high status, we first created a document-term matrix (DTM) of all "informative" words in our corpus of texts and the frequency with which they occurred in each text record (see supplementary information for details). We then fit an elasticnet logistic regression model (with lasso penalty) of *High status* (0/1) as a function of the frequencies of all `r ncol(leader_dtm)-1` words. Words that were strong positive predictors epitomized the semantic content of the text records which provided evidence of high status as a quality of leaders. These included *`r rev(highstatus_plot$data$Word)[1]`* and *`r rev(highstatus_plot$data$Word)[2]`*, which were the strongest predictive terms; *`r rev(highstatus_plot$data$Word)[3]`* and *`r rev(highstatus_plot$data$Word)[4]`* were also relatively strong predictors and *`r rev(highstatus_plot$data$Word)[5]`* and *`r rev(highstatus_plot$data$Word)[6]`*, which were also positive predictors. Notable terms which offered predictive value but with near-zero coefficients included, *influence*, *authority*, *headman*, and *chief*. See Supplementary Figure \@ref(fig:highstatusplot).

@@

@@

The leader quality cluster analysis (Figure \@ref(fig:cluster)A) identified two top-level clusters, one that comprised mostly variables related to prosocial qualities and one that comprised other individual-level qualities which were not necessarily prosocial. Within the latter top-level cluster we identifed two interpretable and strongly supported subclusters which are annotated with text (*Social, reproductive, and material success* and *Competencies*). Within the first top-level cluster, we also identifed two subclusters, one of which contained meansures related to group-level pressures (*Cultural conformity*) and one which contined the prosocial traits (*Prosocial competencies*); these subclusters were moderately supported.

The leader functions cluster analysis (Figure \@ref(fig:cluster)B) also identified two top-level clusters; the larger of which we identified as a management functions cluster (i.e., *Organize cooperation*, *Military command*, *Resolve conflict* functions) and the smaller of which we identified as a prosociality cluster (i.e., *Social functions*, *Distribute resources*, *Hospitality* functions). Within the management cluster, we identified two moderately suppported subclusters; a *Mediation* subcluster (e.g., *Resolve conflict*, *Provide counsel*), and an *Organization* subcluster (e.g., *Organize cooperation*, *Military command*).

@@

Exploratory modeling revealed variation in the vast majority (INCLUDE NO. or %) of our researcher coded variables (TABLEs) did not vary by group type or subsistence strategy (ajusting for effects of leader sex and continental region and the hierarchical structure of ethnographer within culture), suggesting that many of these dimensions of leadership are potnetially universal features of human social groups or at least not strongly impacted by the within or between-group measures investigated here.

@@

Figure \@ref(fig:grouptype) is consistent with a negative relationships between political leadership at the community leadership or above and leadership in kin groups or residential groups. One possible interpretation is that as political structures become more institutionalized and economic systems transcend the household, leadership within households and residential groups diminishes. It is interesting, however, that even among the agricultural populations that are characterized by leadership systems beyond the community level, leadership within residential and family group persists. For example, among rural Korean populations a chairwomen of local organizations maintain important leadership positions within the groups and represent the group to governmental forces. @aa01-062 [p. 173] describes that in these local 'clubs', "the procedure of discussion and decision-making is not based on an idea of democracy but rather on that of unanimity. In this context, the ‘leader’ does not have to be a person of marked influence in the decision-making process. She is not a figurehead but rather a middleman in the meeting, functioning, in effect, as a go-between for the group. There is no despotic idea in the context of a total consensus in this essentially egalitarian society. The chairwoman stands as a middleman between the government and the village. Although the government may call her a ‘leader’ in the village, she properly does not occupy this role."

@@

Our analysis identified several leader functions and qualities that were near cultural universals, and others that were relatively rare. See Figure \@ref(fig:varsupport).

@@

Across a broad range of cultures and social contexts leaders are nearly universally high status males who are described as knowledgeable or intelligent and who are experienced and have a demonstrated history of previous accomplishments or special skills. They most frequently resolve conflicts within their respective groups while also providing important social functions, organizing cooperation among followers, providing counsel or input in followers' affairs, and distributing resources. @oc06-034 [p. 236] in describing kin group leaders among the Iban, small-scale highland horticulturalists primarily in Borneo and Sarawak, provides example of this suite of traits in leaders (which also include elements of prestige-based and charismatic leadership discussed above):

@@

See Figure \@ref(fig:varsupport). These qualities and functions correspond to those highlighted by several evolutionary theories of leadership which have drawn heavily on select ethnographic cases [@garfield_review_2019]. See the *Synthesis* section below.

@@

Several functions and qualities that are theoretically important were relatively rare in the ethnographic record. In the literature on leadership among non-human animals, for example, leaders commonly initiate group movement [@couzin_effective_2005], and "movement" was one of four leadership dimensions that @smith_leadership_2016 identified as comparable among humans and non-human animals, yet movement or migration was identified as a leader function in only `r movement_pct`% of cultures, of which the vast majority (71%) were hunter-gatherers. Among the Blackfoot in the North American plains for example, @nf06-061 [p. 411] describes that, "If there was a dispute as to whether the camp should move north or south, the chief might present his arguments, gain support, and then say that he was going south. He did not order the others to follow, but he knew that they probably would go with him."

Several theories of leadership also emphasize the moral authority of leaders [e.g., @henrich_big_2015; @henrich_evolution_2001; @richerson_tribal_2009; @tietjen_moral_1985; @boggild_evolved_2016], yet leaders were seen as sources of moral authority or "fair" in only a minority of cultures (`r moral_pct`% and `r fairness_pct`% of cultures, respectively). The *Dominance* theory of leadership [e.g., @chapais_competence_2015; @cheng_dominance_2019; @cheng_pride_2010; @henrich_evolution_2001; @tiger_imperial_1971] emphasizes leaders' ability to instill fear, but in only `r feared_pct`% of cultures were leaders characterized as feared. Finally, a large literature on leadership in industrialized societies highlights the attractiveness of leaders [e.g., @sinclair_sexuality_1995; @altemeyer_sexual_1974; @march_leadership_2009], but leaders were characterized as attractive in only `r attractive_pct`% of cultures.

@@

Relationships between leadership and high status, including dependencies and causality are puzzling and currently the focus of much work in evolutionary leadership studies [e.g., @van_vugt_dual_2019; @smith_leadership_2020; @cheng_why_2020; @cheng_dominance_2019; @von_rueden_dynamics_2019]. Drawing on a comparative-evolutionary perspective, incorporating findings from nonhuman social species, @van_vugt_dual_2019 emphasize that although leadership and high social status do indeed often covary, leaders are not necessarily high-status individuals, exemplified when low-ranking individuals in nonhuman species emerge as leaders of group movement or when younger and lower status men and women initiate cooperative activities (such as raids in tribal warfare or social-political movements). @cheng_dominance_2019 and @cheng_why_2020, emphasize, however, that social hierarchies form principally in response to inter-individual variation in status, which may be gained via distinct pathways. Thus, according to @cheng_why_2020, although all leaders are not necessarily high status, especially across nonhuman species, among humans high social status may be a necessary but insufficient condition for achieving and maintaing positions of leadership. In our view, social context is critical to these discussions, as is often mentioned but also under-emphasized and not well supported empirically. Processes of leadership and followership emerge across levels of social organization and individuals simultaneously maintain multiple social statuses. The "low status" male who initiates a inter-group raid on a neighboring group may not be widely respected within his community nor be a candidate for community leadership, however, he might be of exceptionally high status among his peers and high ranking within what @garfield_review_2019 [p. 71] called the "intragenerational social hierarchy."

Our results offer some insight into relationships between leadership and social status across cultures and contexts. As mentioned, high status is the most common quality of leaders (Figure \@ref(fig:varsupport)) and the second most common benefit of leadership and the loss of status the most frequent cost (Figure \@ref(fig:varsupportcb)). 
<!-- mentioned in this sample of the ethnographic record which includes discussions of chiefs, work group leaders, shamans, and age-sets, from a culturally diverse sample including foraging groups and kingdoms.  -->
This lends support to the notion of high social status often being a necessary condition of leader emergence [@cheng_why_2020]. In predicting the ethnographic texts which described leaders as high status, leaders of kin groups and leaders among hunter-gatherers and pastoralists were identified as more likely to be high status relative to some other group and subsistence types. Although discussions of female leaders were rare, the presence of female leaders were also associated with high social status (see Figure \@ref(fig:groupstatus)). Interestingly, political leaders were less likely to be described as high status, which could be related to institutionalized and overt or ascribed social distinctions. Nonetheless, these findings question the informed speculation by @smith_leadership_2020 that, "leadership may therefore be more strongly correlated with high status in large, complex organizations, such as in corporations and governments (p. 2)." However, our text analytic results, which relied on the semantic context of text records to predict the presence of high status leaders generally identified terms related to status and prestige (i.e., *respect*, *prestige*, *honor*) and possibly dominance or proscribed hierarchy (i.e., *rank*, *position*), as stronger predictors of high status leader than terms associated with leadership per se (i.e., *influence*, *authority*, *headman*), which lends some support to the distinction between status and leadership highlighted by @smith_leadership_2020 (though we caution from drawing strong interpretations from these text-analytic results, see Figure \@ref(fig:highstatusplot)). Overall, these results suggest associations between leadership and high status are ubiquitous across human societies, including among kin groups and among hunter-gatherers and that while effective leaders across a wide range of social contexts are the recipients of increased social status for their valuable services, ineffective leadership is associated with reduced social status.

@@

The Competencies subcluster contained a grouping of variables associated with dominance-based leadership, including *Coercive authority* and *Aggressiveness* [@tiger_imperial_1971; @barkow_prestige_1975; @barkow_darwin_1989; @henrich_evolution_2001; @cheng_two_2013] as well as a grouping that included *Knowledgeable/intelligent* and *Experienced/accomplished*, variables that have been emphasized by the *Prestige* theory [@henrich_evolution_2001; @cheng_two_2013] and the more general *Computational services* theory [@hagen_leadership_2019; @garfield_evolutionary_2019], and which are among the most strongly supported in the ethnographic record (Figure \@ref(fig:varsupport)), a point to which we shall return. Interestingly, *Feared* and *Supernatural* clustered more closely with the prestige-style traits than with the dominance-style traits. There is likely a strong connection between shamanism and ritual-based influence and both prestige and dominance styles of leadership. Ritual and supernatural capacities may represent a dimension of leader qualities in which dominance and prestige are coupled forming an alternative style of leadership. Artistic performance is also implicated in this style of leadership and ceremonial, artistic performances by shamans are common displays which often incorporate the superhuman abilities of shamans [@singh_cultural_2017].

@@

@@

multi_aic %>% 
  select(
    var,
    Variable, 
    Sum, 
    uniAIC, 
    multiAIC, 
    AIC_diff,
    pvalue_subsis,
    pvalue_region,
    pvalue_female,
    pvalue_groups
    ) %>% 
  mutate(
    var = map_chr(var, ~ str_split(., '_')[[1]][[1]])
  ) %>% 
  arrange(Variable) %>% 
  kable(
  caption = 'Retained logistic mixed effects models of each variable as functions of subsistence strategy, continental region, leader sex, and group stuctures. Models were retained due to improvement in AIC values relative to an intercept-only model. Sum: the number of rows with evidence for the variable. uniAIC: AIC of intercept-only model. multiAIC: AIC of multivariate model. AIC_diff: difference in AIC values. P-values are the unadjusted p-values of each explanatory variable in the model.',
  digits = 2
  )
```

>>>>

<!-- MALE BIAS -->
Substantial evidence suggests there is a near universal male-bias in community, political, or group-level leadership across human societies [@low_sex_1992; @whyte_status_1978; @sanday_female_1981], as well as across diverse nonhuman social species [@smith_obstacles_2018; @tiger_female_1978]. @low_sex_1992 demonstrated in a systematic cross-cultural sample that political leaders of both sexes were present in only seven of 93 societies and in only two societies were female political leaders identified as equally powerful as their male counterparts. Even among these rare cases, powerful female political leaders are still infrequent within the group and women were more likely to hold influential leadership positions within kin groups [@low_sex_1992]. Empirical ethnographic studies have suggested sex-differences in political influence and leadership across human societies including among more egalitarian cultural contexts are more likely driven by ecological and social structural factors, rather than sex (or gender identity) per se, including social capital and access to same-sex allies [@yanca_female_2004] and/or sex differences in the sexual division of labor [@von_rueden_sex_2018]. At least two studies relying on ethnographic fieldwork comparing leaders and non-leaders in relatively small-scale societies revealed sex is not a strong predictor of community influence or elected leadership positions, after accounting for a broad range of demographic, phenotypic, cognitive, social, and other traits derived from the evolutionary theoretical literature [e.g., @garfield_investigating_2019; @von_rueden_sex_2018].

@@

Evolutionary theoretical models and empirical studies often seek to integrate findings within a broad framework of human cultural diversity. The universality of human leadership systems and the incredible diversity among the qualities of leaders across cultures and contexts and the functions leaders serve is difficult to capture. Scholars are then left to haphazardly self-select ethnographic cases or rely on the self-selected reviews of others for generalizations [e.g., @garfield_review_2019; @von_rueden_leadership_2015]. Here, we seek to provide the first empirical and near comprehensive view of the qualities and functions of leaders across human societies from a representative cross-cultural sample of nonindustrial, non-WEIRD populations. We also provide the first comparative view describing the costs and benefits of leadership for both leaders and followers in an evolutionary biological framework. We also aim to explore how the qualities and functions of leaders vary within populations across social contexts and across populations with different subsistence strategies.

@@

The frequency of leader and follower costs and benefits were estimated similarly to the leader qualities and functions described above. In these data leaders most commonly received benefits in the form of greater access to material resources (documented in about 70% of cultures) and increased social status (documented in about 65% of cultures). Mating benefits and receiving social services were also relatively frequent benefits of leadership. In general, the costs tracked the benefits. Leaders were often described as losing their high social status (documented in about 45% of cultures) and greater access to resources (documented in about 40% of cultures). See Figure \@ref(fig:varsupportcb).

Followers on the other hand most commonly benefited from leadership systems through receiving material resources (documented in about 50% of cultures) but otherwise most frequently through social services provided (documented in about 35% of cultures). Mating and fitness benefits were relatively rare for followers. Followers were most frequently described as incurring costs in the form of material resources lost (documented in about 40% of cultures) and in providing social services (documented in about 25% of cultures). See Figure \@ref(fig:varsupportcb).

@@

For subsistence strategy, hunter-gatherers had relatively more evidence for *Polygyny*, *Leader-Mating benefits*, *No coercive authority*, *Political appointments* and *Military command* than other subsistence strategies. Pastoralists had more evidence for *Ritual functions*, *Supernatural*, *Aggressiveness* and *High quality spouse*. Hunter-gatherers and pastoralists both had more evidence for *High status*, *Providing counsel/direction*, and *Providing subsistence*. The remaining subsistence strategies generally had relatively less evidence for all variables, except that the mixed strategy had relatively more evidence for *Organize cooperation*, *Distribute resources*, and *Prosocial investment*. See Figure S\@ref(fig:heatmapsubsis).

@@

There was strong patterning of relative evidence for almost all variables by leader sex. Female leaders had relatively more evidence for *High quality spouse*, *Leader - Mating benefits*, *Polygyny*, *High status*, and *Age*. Male leaders had relatively more evidence for *Follower - Social  services benefits*, *Political appointments*, *Prosocial investment*, and *Leader-Social services costs*. Records pertaining to both sexes had relatively more evidence for *Organize cooperation*, *Providing subsistence*, *Ritual functions*, *Bravery*, *Experience/accomplished*, and few that overlapped with males, such as *Military command* and *Distributing resources*. See Figure S\@ref(fig:heatmapsex).

There was also strong patterning of relative evidence for almost all variables by group structure. There was relatively more evidence for military traits in military groups, relatively more evidence for ritual and supernatural traits in religious groups, relatively more evidence for organizing cooperation and providing subsistence in economic groups, and so forth. Kin groups had relatively more evidence for *Provide counsel/direction*, *Prosocial investment*, *Age*, and *High status*. Community and residential groups had relatively low evidence for most variables, except that there was relatively more evidence for *High quality spouses* among residential group leaders. See Figure S\@ref(fig:heatmapgroups).

@@

We then explored relationships between total leader benefits and costs, summed across all leader benefit and cost variables, respectively, by the predictor variables used above (subsistence strategy, continental region, leader sex, and group structure). There was less evidence for leader costs, however, which limited our ability to fit exploratory regression models. Therefore, in modeling total leader costs, we used subsistence and group structure as predictors. In general, leaders at the highest supracommunity political level had the most evidence for receiving benefits and the least evidence for suffering costs and leaders of military groups has the highest total costs. See Figure \@ref(fig:leadertotalbenefits).

@@

*High status* was the most frequently identified leader quality in the sample, with `r statusPercTxt`% of text records and `r statusPercCult`% of cultures containing evidence that leaders had high social status. To explore factors that distinguished these records from those with no evidence for high status, we first created a document-term matrix (DTM) of all "informative" words in our corpus of texts and the frequency with which they occurred in each text record (see supplementary information for details). We then fit an elasticnet logistic regression model (with lasso penalty) of *High status* (0/1) as a function of the frequencies of all `r ncol(leader_dtm)-1` words. Words that were strong positive predictors epitomized the semantic content of the text records which provided evidence of high status as a quality of leaders. These included *`r rev(highstatus_plot$data$Word)[1]`* and *`r rev(highstatus_plot$data$Word)[2]`*, which were the strongest predictive terms; *`r rev(highstatus_plot$data$Word)[3]`* and *`r rev(highstatus_plot$data$Word)[4]`* were also relatively strong predictors; and *`r rev(highstatus_plot$data$Word)[5]`* and *`r rev(highstatus_plot$data$Word)[6]`*, which were also positive predictors. Notable terms which offered predictive value but with near-zero coefficients included, *influence*, *authority*, *headman*, and *chief*. See Figure S\@ref(fig:highstatusplot).

@@

<!-- NEED TO DO TABLE -->

@@

The leader quality cluster analysis (Figure \@ref(fig:cluster)A) identified two top-level clusters, one that comprised mostly variables related to prosocial qualities and one that comprised other individual-level qualities which were not necessarily prosocial. Within the latter top-level cluster we identified two interpretable and strongly supported subclusters which are annotated with text (*Social, reproductive, and material success* and *Competencies*). Within the first top-level cluster, we also identified two subclusters, one of which contained measures related to group-level pressures (*Cultural conformity*) and one which contained the prosocial traits (*Prosocial competencies*); these subclusters were moderately supported.

The leader functions cluster analysis (Figure \@ref(fig:cluster)B) also identified two top-level clusters; the larger of which we identified as a management functions cluster (e.g., *Organize cooperation*, *Military command*, *Resolve conflict*) and the smaller of which we identified as a prosociality cluster (e.g., *Social functions*, *Distribute resources*, *Hospitality*). Within the management cluster, we identified two moderately supported subclusters; a *Mediation* subcluster (e.g., *Resolve conflict*, *Provide counsel*), and an *Organization* subcluster (e.g., *Organize cooperation*, *Military command*).

@@

Exploratory modeling revealed variation in the vast majority (INCLUDE NO. or %) of our researcher coded variables (TABLEs) did not vary by group type or subsistence strategy (adjusting for effects of leader sex and continental region and the hierarchical structure of ethnographer within culture), suggesting that many of these dimensions of leadership are potentially universal features of human social groups or at least not strongly impacted by the within or between-group measures investigated here.

@@

Figure \@ref(fig:grouptype) is consistent with a negative relationships between political leadership at the community leadership or above and leadership in kin groups or residential groups. One possible interpretation is that as political structures become more institutionalized and economic systems transcend the household, leadership within households and residential groups diminishes. It is interesting, however, that even among the agricultural populations that are characterized by leadership systems beyond the community level, leadership within residential and family group persists. For example, among rural Korean populations a chairwomen of local organizations maintain important leadership positions within the groups and represent the group to governmental forces. @aa01-062 [p. 173] describes that in these local 'clubs', 
>"the procedure of discussion and decision-making is not based on an idea of democracy but rather on that of unanimity. In this context, the ‘leader’ does not have to be a person of marked influence in the decision-making process. She is not a figurehead but rather a middleman in the meeting, functioning, in effect, as a go-between for the group. There is no despotic idea in the context of a total consensus in this essentially egalitarian society. The chairwoman stands as a middleman between the government and the village. Although the government may call her a ‘leader’ in the village, she properly does not occupy this role."

@@

Our analysis identified several leader functions and qualities that were near cultural universals, and others that were more variable or relatively rare. See Figure \@ref(fig:varsupport).

@@

Across a broad range of cultures and social contexts leaders are nearly universally high status males who are described as knowledgeable or intelligent, are experienced, and have a demonstrated history of previous accomplishments or special capacities. They most frequently resolve conflicts within their respective groups while also providing important social functions, organizing cooperation among followers, providing counsel or input in followers' affairs, and distributing resources. @oc06-034 [p. 236] in describing kin group leaders among the Iban, small-scale highland horticulturalists primarily in Borneo and Sarawak, provides example of this suite of traits in leaders (which also include elements of prestige-based and charismatic leadership discussed above):

@@

See Figure \@ref(fig:varsupport). These qualities and functions correspond to those highlighted by several evolutionary theories of leadership which have drawn heavily on select ethnographic cases [@garfield_review_2019]. See the *Synthesis* section below.
<!-- NEED TO DO ADD SECTION OR DELETE HERE -->

@@

Several functions and qualities that are theoretically important were relatively rare in the ethnographic record. In the literature on leadership among nonhuman animals, for example, leaders commonly initiate group movement [@couzin_effective_2005], and "movement" was one of four leadership dimensions that @smith_leadership_2016 identified as comparable among small-scale human societies and nonhuman animal societies, yet movement or migration was identified as a leader function in only `r movement_pct`% of cultures, of which the vast majority (71%) were hunter-gatherers. Among the Blackfoot in the North American plains for example, @nf06-061 [p. 411] describes that, "if there was a dispute as to whether the camp should move north or south, the chief might present his arguments, gain support, and then say that he was going south. He did not order the others to follow, but he knew that they probably would go with him."

Several theories of leadership also emphasize the moral authority of leaders [e.g., @henrich_big_2015; @henrich_evolution_2001; @richerson_tribal_2009; @tietjen_moral_1985; @boggild_evolved_2016], yet leaders were seen as sources of moral authority or "fair" in only a minority of cultures (`r moral_pct`% and `r fairness_pct`% of cultures, respectively). The *Dominance* theory of leadership [e.g., @chapais_competence_2015; @cheng_dominance_2019; @cheng_pride_2010; @henrich_evolution_2001; @tiger_imperial_1971] emphasizes leaders' ability to instill fear, but in only `r feared_pct`% of cultures were leaders characterized as feared. Finally, a large literature on leadership in industrialized societies highlights the physical attractiveness of leaders [e.g., @sinclair_sexuality_1995; @altemeyer_sexual_1974; @march_leadership_2009], but leaders were characterized as attractive in only `r attractive_pct`% of cultures.

@@

Relationships between leadership and high status, including causality and dependence, are puzzling and currently the focus of much work in evolutionary leadership studies [e.g., @van_vugt_dual_2019; @smith_leadership_2020; @cheng_why_2020; @cheng_dominance_2019; @von_rueden_dynamics_2019]. Drawing on a comparative-evolutionary perspective, incorporating findings from nonhuman social species, @van_vugt_dual_2019 emphasize that although leadership and high social status do indeed often covary, leaders are not necessarily high-status individuals, exemplified when low-ranking individuals in nonhuman species emerge as leaders of group movement or when younger and lower status men and women initiate cooperative activities (such as raids in small-scale warfare or social-political movements). @cheng_dominance_2019 and @cheng_why_2020, emphasize, however, that social hierarchies form principally in response to inter-individual variation in status, which may be gained via distinct pathways. Thus, according to @cheng_why_2020, although all leaders are not necessarily high status, especially across nonhuman species, among humans high social status may be a necessary but insufficient condition for achieving and maintaining positions of leadership. In our view, social context is critical to these discussions, as is often mentioned but also overlooked. Processes of leadership and followership emerge across levels of social organization and individuals simultaneously maintain multiple social statuses. The "low status" male who initiates an inter-group raid on a neighboring group may not be widely respected within his community nor be a candidate for community leadership, however, he might be of exceptionally high status among his peers and high ranking within what @garfield_review_2019 [p. 71] called the "intragenerational social hierarchy."

Our results offer some insight into relationships between leadership and social status across cultures and contexts. As mentioned, high status is the most common quality of leaders (Figure \@ref(fig:varsupport)) and the second most common benefit of leadership and the loss of status the most frequent cost (Figure \@ref(fig:varsupportcb)). 
<!-- mentioned in this sample of the ethnographic record which includes discussions of chiefs, work group leaders, shamans, and age-sets, from a culturally diverse sample including foraging groups and kingdoms.  -->
This lends support to the notion of high social status often being a necessary condition of leader emergence [@cheng_why_2020]. In predicting the ethnographic texts which described leaders as high status, leaders of kin groups and leaders among hunter-gatherers and pastoralists were identified as more likely to be high status relative to some other group and subsistence types. Although discussions of female leaders were rare, the presence of female leaders were also associated with high social status (see Figure \@ref(fig:groupstatus)). Interestingly, political leaders were less likely to be described as high status, which could be related to institutionalized and overt or ascribed social distinctions. Nonetheless, these findings question the informed speculation by @smith_leadership_2020 that, "leadership may therefore be more strongly correlated with high status in large, complex organizations, such as in corporations and governments (p. 2)." However, our text analytic results, which relied on the semantic context of text records to predict the presence of high status leaders generally identified terms related to status and prestige (i.e., *respect*, *prestige*, *honor*) and possibly dominance or proscribed hierarchy (i.e., *rank*, *position*), as stronger predictors of high status leader than terms associated with leadership per se (i.e., *influence*, *authority*, *headman*), which lends some support to the distinction between status and leadership highlighted by @smith_leadership_2020 (though we caution from drawing strong interpretations from these text-analytic results, see Figure S\@ref(fig:highstatusplot)). Overall, these results suggest associations between leadership and high status are ubiquitous across human societies, including among kin groups and among hunter-gatherers and that while effective leaders across a wide range of social contexts are the recipients of increased social status for their valuable services, ineffective leadership is associated with reduced social status.

@@

The Competencies subcluster contained a grouping of variables associated with dominance-based leadership, including *Coercive authority* and *Aggressiveness* [@tiger_imperial_1971; @barkow_prestige_1975; @barkow_darwin_1989; @henrich_evolution_2001; @cheng_two_2013] as well as a grouping that included *Knowledgeable/intelligent* and *Experienced/accomplished*, variables that have been emphasized by the *Prestige* theory [@henrich_evolution_2001; @cheng_two_2013] and the more general *Computational services* theory [@hagen_leadership_2019; @garfield_evolutionary_2019], and which are among the most strongly supported in the ethnographic record (Figure \@ref(fig:varsupport)), a point to which we shall return. Interestingly, *Feared* and *Supernatural* clustered more closely with the prestige-style traits than with the dominance-style traits. There is likely a strong connection between shamanism and ritual-based influence with both prestige and dominance styles of leadership. Ritual and supernatural capacities may represent a dimension of leader qualities in which dominance and prestige are coupled forming an alternative style of leadership, as shamans often provide a necessary and valued public good, have special expertise and abilities, are feared and achieve or maintain influence via threats or use of direct and/or indirect aggression. Artistic performance is also implicated in this style of leadership and ceremonial, artistic performances by shamans are common displays which often incorporate the superhuman abilities of shamans [@singh_cultural_2017].

@@

## Female leadership and sex differences

Although only a small fraction of the ethnographic record sampled here discussed female leaders, our results do provide some insight on the nature of female leadership across human societies and sex-differences in leadership.

Female leaders were more likely to emerge in kin groups and residential subgroups, consistent with theoretical arguments emphasizing women as leaders of families by @garfield_evolutionary_2019 and @hagen_leadership_2019. Texts describing female leaders exclusively also provided a relatively higher proportion of total leader benefits (Figure \@ref(fig:leadertotalbenefits) and descriptions of high status leaders (Figure \@ref(fig:groupstatus)) compared to descriptions of leaders that were not sex-specific (holding other measures constant; although this effect was similar to benefits to and high status of male-leader texts). This suggests, much like male leaders who are more often political leaders, female leaders are also high status individuals and receive substantial benefits for their services.

Across cultures and contexts, when ethnographic texts described female leaders exclusively, they were relatively more likely to be described as older, of high status, married to a high-quality spouse in a polygynous marriage, and the recipients of various mating benefits. This suite of characteristics is consistent with discussions of high status first-wives who gain and maintain increased social influence across the lifespan by leveraging their extended kin and social networks [@goodale_tiwi_1971; @brown_her_1985], and supports the view of positive assortative mating of high status leaders @garfield_investigating_2019 documented among elected leaders in a small-scale society with relative gender-egalitarianism. Evidence for male leaders (relative to evidence for female leaders or evidence that was not sex-specific) was indeed linked to group-level and political functions and male leaders were relativey more likely to provide social services, control political structure, and were at greater risk of losing their high social standing. See Figure S\@ref(fig:heatmapsex).

Female leaders were also more likely to lack coercive authority, which is consistent with much of leadership studies suggesting dominant and aggressive women are less likely to be preferred as leaders, unlike dominent men who are sometimes prefered as leaders [@garfield_investigating_2019; @williams_subtle_2016].

Despite sex differences, there is also likely to be substantial overlap in many of the qualities and functions of female and male leaders across human societies. Leader sex was not a predictor of the majority of the measures investigated here (adjusting for other population-level measures; see Table S\@ref(tab:universalvars)). When sex was a predictor (Table S\@ref(tab:variablevars)), descriptions including both female and male leaders provided relatively more supporting evidence for many key measures of leader qualities and functions, including group organization, resource distribution, ritual functions, and being brave and experienced and accomplished.

@@

multi_aic %>% 
  select(
    var,
    Variable, 
    Sum, 
    uniAIC, 
    multiAIC, 
    AIC_diff,
    pvalue_subsis,
    pvalue_region,
    pvalue_female,
    pvalue_groups
    ) %>% 
  mutate(
    var = map_chr(var, ~ str_split(., '_')[[1]][[1]])
  ) %>% 
  arrange(Variable) %>% 
  kable(
  caption = 'Retained logistic mixed effects models of each variable as functions of subsistence strategy, continental region, leader sex, and group structures. Models were retained due to improvement in AIC values relative to an intercept-only model. Sum: the number of rows with evidence for the variable. uniAIC: AIC of intercept-only model. multiAIC: AIC of multivariate model. AIC_diff: difference in AIC values. P-values are the unadjusted p-values of each explanatory variable in the model.',
  digits = 2
  )
```
